Split-Apply-Combine with Collapsing Groups

### 2022-12-29 paper/article.Rnw
some writing, moved to jss format

\documentclass[11pt, fleqn]{article}
\usepackage[english]{babel}
\usepackage{amsmath, amssymb}

\usepackage{colortbl, xcolor}

\usepackage[ruled, linesnumbered]{algorithm2e}

\usepackage{natbib}
\setlength{\parskip}{\baselineskip}%
\setlength{\parindent}{0pt}%

\title{Spit-Apply-Combine with Collapsing Groups}
\author{Mark P.J. van der Loo}
\date{CONCEPT \today{}}

\begin{document}
\maketitle

\section{Introduction}
The operation of splitting a data set into non-overlapping groups, computing an
aggregate for each group, and combining the results into a new datasets is one
of the most common operations in data analyses. The idea behind this so-called
split-apply-combine approach is that the calculation for each groups only uses
data available within the group. However, in some cases the quality of data in
a group may be insufficient for producing a reliable group aggregate. Common
quality issues include insufficient records, or too many records with values
for certain variables missing. In such cases it is not uncommon to use data
from outside the target group  to estimate the group aggregate. One example is
Small Area Estimation, where the lack of observations in one group is
compensated by using data from similar but other groups. Another area where
such approaches occur is group-wise imputation. For example, one may wish to
impute a group estimate (say, mean, or median) for missing values. If there are
too little observations to provide an accurate estimate of the group mean, the
group may be expanded, trading bias for variance.

@@

In this paper we formalise the concept of `collapsing scheme', and give an
algorithm for split-apply-combine operations based on it. Next, we demonstrate
the R package \texttt{rcollapse} that implements this algorithm together with
some utility functions that make working with collapsing schemes easier.

@@

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ stands for `the set of all
subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
$f$. That is: you give it a set of labels from $A$ and it returns all the
elements of $U$ that have any of those labels. This definition is slightly too
general for the current algorithm, but it will come in handy when we move to
collapsing groups.

@@

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{equation}
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{equation}
Here, $A_0$ is the most detailed partition, splitting a dataset into seven
groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
$0113$ and the second, labeled $01.2$ is constructed by combining
$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.

@@

\section{The \texttt{rcollapse} package}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  collapse(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

To accommodate the different use cases, the \texttt{rcollapse} package supports
two different interfaces for specifying collapse sequences.  The first and most
general is the formula interface, that requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface discussed before. This only supports a single grouping
variable. For hierarchical classifications there is a helper function to
translate parent-child representations of such classifications into a
collapsing sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables \texttt{A} and
\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
variable \texttt{B1} and if that is not sufficient, in the next step grouping
in the second variable is dropped so grouping is only defined by \texttt{A}.
In \texttt{rcollapse} such a scheme is denoted with the formula \texttt{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \texttt{+}.

@@

In R code the previous example can be executed as follows.
\begin{verbatim}
  > library(rcollapse)
  > input <- data.frame(
  +      A  = c(1,1,1,2,2,2,3,3,3)
  +    , B  = c(11,11,11,12,12,13,21,22,12)
  +    , B1 = c(1,1,1,1,1,1,2,2,1)
  +    , Y  = 2^(0:8)
  + )
  >
  > collapse(data     = input
  +        , collapse = A*B ~ A*B1 + A
  +        , test     = function(d) nrow(d)>=3
  +        , tY       = sum(Y) )
    A  B level  tY
  1 1 11     0   7
  2 2 12     1  56
  3 2 13     1  56
  4 3 21     2 448
  5 3 22     2 448
  6 3 12     2 448
\end{verbatim}
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame, which may require some extensive data manipulation.

@@

\subsection{Programming over \texttt{collapse}}

@@

\bibliographystyle{plainnat}
\bibliography{vanderloo}
\end{document}

>>>>

\documentclass[article,table]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------
%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage{amsmath, amssymb}
\usepackage[ruled, linesnumbered]{algorithm2e}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
        Statistics Netherlands and University of Leiden}

\Plainauthor{Mark P.J. van der Loo}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Split-Apply-Combine with Collapsing Groups}
\Plaintitle{Split-Apply-Combine with Collapsing Groups}
\Shorttitle{Split-Apply-Combine with Collapsing Groups}

\Abstract{
foo
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{data analysis, \proglang{R}}
\Plainkeywords{data analysis, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
  Research and Development\\
  Statistics Netherlands\\
  Henri Faasdreef 312\\
  2492JP Den Haag, the Netherlands\\
  E-mail: \email{mpj.vanderloo@cbs.nl}\\
  URL: \url{https://www.markvanderloo.eu}\\
  \emph{and}\\
  Leiden Institute of Advanced Computer Science (LIACS)\\
  P.O. Box 9512\\
  2300 RA Leiden, The Netherlands\\
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.

\section{Introduction}
The operation of splitting a data set into non-overlapping groups, computing an
aggregate for each group, and combining the results into a new dataset is one
of the most common operations in data analyses. Indeed, any software for data
analyses includes some functionality for this. For example, the combination of
\code{split}/\code{lapply}/\code{unsplit} as well as \code{aggregate} have been
a part of the \proglang{S} \citep{becker1988new} and \proglang{R} \citep{rcore}
languages for a long time. For \proglang{R} there are several packages that
implement functionality for this, including \pkg{plyr}
\citep{wickham2011split}, it's successor \pkg{dplyr} \citep{wickham2022dplyr},
it's drop-in replacement \pkg{poorman} \citep{eastwood2022poorman}, and
\proglang{R} packages \pkg{collapse} \citep{krantz2022collapse} and
\pkg{data.table} \citep{dowle2022datatable}. In \proglang{Python} the
\code{pandas} package implements several methods for grouping records and
aggregating over one or more columns in data frame objects. Similarly, the
\pkg{DataFrames} package for \proglang{julia} implements such functionality
\citep{kaminski2022dataframes}.

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. Small area estimation is a collection of
bias-variance trade-offs where one `borrows statistical strength' from
out-of-group records to obtain a stable estimate for groups with few
observations. The out-of-group records can be obtained, for example by
combining the original group with a group of records that are deemed similar in
certain respects. A second area where out-of-group records play a role is in
certain hot-deck imputation methods \cite{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group. In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. A third example comes from the field of statistical disclosure control
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule states
that the value in a grouped aggregation may not be determined for more than $p$
percent by a single entity. This occurs for example, when a certain economic
activity in a geographic region is dominated by a single company. In such cases
the value can either not be published, or one publishes an aggregate over
larger groups.

Common quality issues
include insufficient records, or too many records with values for certain
variables missing. In such cases it is not uncommon to use data from outside
the target group  to estimate the group aggregate. One example is Small Area
Estimation, where the lack of observations in one group is compensated by using
data from similar but other groups. Another area where such approaches occur is
group-wise imputation. For example, one may wish to impute a group estimate
(say, mean, or median) for missing values. If there are too little observations
to provide an accurate estimate of the group mean, the group may be expanded,
trading bias for variance.

@@

In this paper we formalise the concept of `collapsing scheme', and give an
algorithm for split-apply-combine operations based on it. Next, we demonstrate
the R package \texttt{accumulate} that implements this algorithm together with
some utility functions that make working with collapsing schemes easier.

@@

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}

Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ stands for `the set of all
subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
$f$. That is: you give it a set of labels from $A$ and it returns all the
elements of $U$ that have any of those labels. This definition is slightly too
general for the current algorithm, but it will come in handy when we move to
collapsing groups.

@@

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{table}[h!]
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{table}

Here, $A_0$ is the most detailed partition, splitting a dataset into seven
groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
$0113$ and the second, labeled $01.2$ is constructed by combining
$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.

@@

\section{The accumulate R package}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  accumulate(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

To accommodate the different use cases, the \pkg{accumulate} package supports
two different interfaces for specifying collapse sequences.  The first and most
general is the formula interface, that requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface discussed before. This only supports a single grouping
variable. For hierarchical classifications there is a helper function to
translate parent-child representations of such classifications into a
collapsing sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables \texttt{A} and
\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
variable \texttt{B1} and if that is not sufficient, in the next step grouping
in the second variable is dropped so grouping is only defined by \texttt{A}.
In \pkg{accumulate} such a scheme is denoted with the formula \texttt{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \texttt{+}.

@@

In R code the previous example can be executed as follows.
\begin{verbatim}
  > library(accumulate)
  > input <- data.frame(
  +      A  = c(1,1,1,2,2,2,3,3,3)
  +    , B  = c(11,11,11,12,12,13,21,22,12)
  +    , B1 = c(1,1,1,1,1,1,2,2,1)
  +    , Y  = 2^(0:8)
  + )
  >
  > collapse(data     = input
  +        , collapse = A*B ~ A*B1 + A
  +        , test     = function(d) nrow(d)>=3
  +        , tY       = sum(Y) )
    A  B level  tY
  1 1 11     0   7
  2 2 12     1  56
  3 2 13     1  56
  4 3 21     2 448
  5 3 22     2 448
  6 3 12     2 448
\end{verbatim}
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame, which may require some extensive data manipulation.

@@

\subsection{Programming over accumulate }

@@

\bibliography{refs}

\end{document}
